### 2021-10-18 text/outline/outline.tex
finish section 9.1.1

% Several applied methods for linking perturbations to changes in gene expression in single-cell CRISPR screens ; however, these methods focus on hypothesis test (rather than estimation) 
% SCEPTRE \cite{Barry2020}

@@

We study the thresholding method from a theoretical perspective in an idealized Gaussian setting. Suppose we observe data $(g_1, m_1), \dots, (g_n, m_n)$ from the following model:

\begin{equation}\label{theoretical_model}
\begin{cases}
m_i = \beta^m_0 + \beta^m_1 p_i + \ep_i \\
g_i = \beta^g_0 + \beta^g_1 p_i + \tau_i \\
p_i \sim \textrm{Bern}(\pi) \\
\ep_i, \tau_i \sim N(0,1) \\
p_i \indep \tau_i \indep \ep_i.
\end{cases}
\end{equation}
For a given threshold $c \in \R$, let $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_0$ for $\beta^m_0$ is given by $$\hat{\beta}^m_0 = \frac{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}}) (m_i - \overline{m})}{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})}.$$ The limit in probability (as $n \to \infty$) of $\hat{\beta}^m_0$ is
$$\hat{\beta}^m_1 \xrightarrow{P} \beta^m_1 \left(\frac{ \pi( \omega - \E[ \hat{p}_i ])}{ \E[\hat{p}_i] (1 - \E[\hat{p}_i])}\right),$$ where
$$\begin{cases}
\omega = \pi\left( 1 - \Phi(c - \beta_1^g - \beta_0^g) \right),\\ \zeta = (1-\pi)(1 - \Phi\left( c - \beta^g_0 \right)), \\ \E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi. 
\end{cases}$$
Let $\gamma: \R^4 \to \R$ be defined by
$$ \gamma( \beta^g_0, \beta^g_1, c, \pi ) = \frac{\pi (\omega - \E[\hat{p}_i])}{\E[\hat{p}_i] (1 - \E[\hat{p}_i]) }.$$ We call $\gamma$ the \textit{attenuation function}. Observe that \begin{itemize}
\item[i.] $\gamma$ does not depend on $\beta^m_1$ or $\beta^m_0$, and
\item[ii.] $\hat{\beta}^m_1 \xrightarrow{P} [\gamma(\beta_0^g, \beta_1^g, c, \pi)] \beta^m_1.$
\end{itemize}
Let $b: \R^4 \to \R$ be the asymptotic bias of $\hat{\beta}^m_1$. We can express $b$ in terms of $\gamma$ as $$b\left(\beta^g_0, \beta^g_1, c, \pi \right) = \beta^m_1 - \gamma( \beta_0^g, \beta_1^g, c, \pi) \beta^m_1 = \beta^m_1 \left[ 1- \gamma(\beta^g_0, \beta^g_1, c, \pi) \right].$$ The asymptotic bias vanishes when the attenuation function equals $1$.

@@

Next, we generalize the expression for bias to arbitrary $\pi \in [0,1/2]$. The following simple relationship holds:
$$ \lim_{ c \to \infty } b(\beta^g_0, \beta^g_1, c, \pi) = \pi.$$ In other words, when the perturbation probability is equal to $\pi$, setting the threshold to a large number yields a bias of $\pi$ (Figure \ref{thresholding_theoretical}b). We can build intuition for this result by considering an extreme example. When $\pi$ is very small (e.g., $\pi = 0.01$), almost all cells are unperturbed. Therefore, in selecting a large threshold, one correctly classifies nearly all \textit{unperturbed} cells as \textit{unperturbed}; on the other hand, the \textit{perturbed} cells that one misclassifies as \textit{unperturbed} are swamped in number by the truly \textit{unperturbed} cells, resulting in a small bias.

@@

Finally, we study the variance of the thresholding estimator. Let $v$ denote the limiting variance of scaled estimator $\sqrt{n} \hat{\beta}^m_1$, i.e.,
$$v(\beta^g_0, \beta^g_1, c, \pi) = \lim_{n \to \infty} \V(\sqrt{n} \hat{\beta}^m_1).$$ The limiting variance of $\sqrt{n} \hat{\beta}^m_1$ grows without bound as the threshold $c$ tends to infinity: 
$$ \lim_{c \to \infty} v(\beta^g_0, \beta^g_1, c, \pi) = \infty.$$ Thus, while selecting a large threshold is effective from a bias reduction standpoint when $\pi$ is small, it causes the variance to explode.

@@

\subsection{Theoretical results for thresholding estimator}

>>>>

@@

We study the thresholding method from a theoretical perspective in an idealized Gaussian setting. Suppose that we observe data $(g_1, m_1), \dots, (g_n, m_n)$ from the following model:

\begin{equation}\label{theoretical_model}
\begin{cases}
m_i = \beta^m_0 + \beta^m_1 p_i + \ep_i \\
g_i = \beta^g_0 + \beta^g_1 p_i + \tau_i \\
p_i \sim \textrm{Bern}(\pi) \\
\ep_i, \tau_i \sim N(0,1) \\
p_i \indep \tau_i \indep \ep_i.
\end{cases}
\end{equation}
For a given threshold $c \in \R$, let $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ for $\beta^m_1$ is given by $$\hat{\beta}^m_1 = \frac{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}}) (m_i - \overline{m})}{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})}.$$ The limit in probability (as $n \to \infty$) of $\hat{\beta}^m_0$ is
\begin{equation}\label{thresh_est_intercepts}
\hat{\beta}^m_1 \xrightarrow{P} \beta^m_1 \left(\frac{ \pi( \omega - \E[ \hat{p}_i ])}{ \E[\hat{p}_i] (1 - \E[\hat{p}_i])}\right),
\end{equation}
where
$$\begin{cases}
\omega = 1 - \Phi\left(c - \beta_1^g - \beta_0^g \right) ,\\ \zeta = 1 - \Phi\left( c - \beta^g_0 \right), \\ \E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi. 
\end{cases}$$
Let $\gamma: \R^4 \to \R$ be defined by
$$ \gamma( \beta^g_0, \beta^g_1, c, \pi ) = \frac{\pi (\omega - \E[\hat{p}_i])}{\E[\hat{p}_i] (1 - \E[\hat{p}_i]) }.$$ We call $\gamma$ the \textit{attenuation function}. Observe that \begin{itemize}
\item[i.] $\gamma$ does not depend on $\beta^m_1$ or $\beta^m_0$, and
\item[ii.] $\hat{\beta}^m_1 \xrightarrow{P} [\gamma(\beta_0^g, \beta_1^g, c, \pi)] \beta^m_1.$
\end{itemize}
Let $b: \R^4 \to \R$ be the asymptotic bias of $\hat{\beta}^m_1$. We can express $b$ in terms of $\gamma$ as $$b\left(\beta^g_0, \beta^g_1, c, \pi \right) = \beta^m_1 - \gamma( \beta_0^g, \beta_1^g, c, \pi) \beta^m_1 = \beta^m_1 \left[ 1- \gamma(\beta^g_0, \beta^g_1, c, \pi) \right].$$ The asymptotic bias vanishes when the attenuation function equals $1$.

@@

Next, we generalize the expression for bias when the threshold is large to arbitrary $\pi \in [0,1/2]$. The following simple relationship holds:
$$ \lim_{ c \to \infty } b(\beta^g_0, \beta^g_1, c, \pi) = \pi.$$ In other words, when the perturbation probability is equal to $\pi$, setting the threshold to a large number yields a bias of $\pi$ (Figure \ref{thresholding_theoretical}b). We can build intuition for this result by considering an extreme example. When $\pi$ is very small (e.g., $\pi = 0.01$), almost all cells are unperturbed. Therefore, in selecting a large threshold, one correctly classifies nearly all unperturbed cells as unperturbed; on the other hand, the \textit{perturbed} cells that one misclassifies as \textit{unperturbed} are swamped in number by the truly unperturbed cells, resulting in a small bias.

@@

Finally, to shed light on limitations of the large threshold selection strategy, we study the variance of the thresholding estimator. Let $v$ denote the limiting variance of scaled estimator $\sqrt{n} \hat{\beta}^m_1$, i.e.,
$$v(\beta^g_0, \beta^g_1, c, \pi) = \lim_{n \to \infty} \V(\sqrt{n} \hat{\beta}^m_1).$$ The limiting variance of $\sqrt{n} \hat{\beta}^m_1$ grows without bound as the threshold $c$ tends to infinity: 
$$ \lim_{c \to \infty} v(\beta^g_0, \beta^g_1, c, \pi) = \infty.$$ Thus, while selecting a large threshold is effective from a bias reduction standpoint when $\pi$ is small, it causes the variance to explode.

@@

\subsection{Theoretical details for thresholding estimator}

This section contains proofs for results presented the section ``Theoretical analysis of thresholding estimator.''

\subsubsection{Limit of $\hat{\beta}^m_1$ in Gaussian model with intercepts}

Consider the thresholding estimator $\hat{\beta}^m_1$ for the Gaussian model with intercepts (\ref{theoretical_model}). Dividing by $n$ in (\ref{thresh_est_intercepts}), we can express $\hat{\beta}^m_1$ as
$$ \hat{\beta}^m_1 = \frac{ \frac{1}{n} \sum_{i=1}^n ( \hat{p}_i - \overline{\hat{p}_i})(m_i - \overline{m})}{ \frac{1}{n} \sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})}.$$ By WLLN,
$$ \hat{\beta}^m_1 \xrightarrow{P} \frac{\textrm{Cov}(\hat{p}_i, m_i)}{\V\left(\hat{p}_i\right)}.$$ To compute this quantity, we first compute several simpler quantities:
\begin{itemize}
\item[1.] \textbf{Expectation of $m_i$}: $\E[m_i] = \beta^m_0 + \beta^m_1\pi.$
\item[2.] \textbf{Expectation of $\hat{p}_i$}: Let $f$ denote the $N(0,1)$ density. Then \begin{multline*}
\E[\hat{p}_i] = \P\left[\hat{p}_i = 1\right] = \P\left[\beta^g_0 + \beta^g_1 + \tau_i \geq c \right] = \\ \textrm{(By LOTP) } \P\left[ \beta^g_0 + \tau_i \geq c \right]\P\left[p_i = 0\right] + \P\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \P[p_i = 1] \\ = \P\left[ \tau_i \geq c - \beta^g_0\right](1- \pi) + \P\left[ \tau_i \geq c - \beta^g_1 - \beta^g_0 \right](\pi) \\ = \left(\int_{c - \beta^g_0}^\infty f\right) (1 - \pi) + \left( \int_{c - \beta^g_1 - \beta^g_0}^\infty f \right)(\pi) = \zeta(1-\pi) + \omega \pi,
\end{multline*}
where $$\zeta = \int_{c - \beta_0^g}^\infty f = 1 - \Phi(c - \beta^g_0)$$ and 
$$ \omega = \int_{c - \beta^g_1 - \beta^g_0} f = 1 - \Phi(c - \beta^g_1 - \beta^g_0).$$
\item[3.] \textbf{Expectation of $\hat{p}_i p_i$}: 
$$\E\left[ \hat{p}_i p_i \right] = \E\left[ \hat{p}_i | p_i = 1 \right] \P\left[ p_i =1 \right] = \P\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \pi = \omega \pi.$$
\item[4.] \textbf{Expectation of $\hat{p}_i m_i$}:
\begin{multline*}
\E\left[\hat{p}_i m_i\right] = \E[\hat{p}_i (\beta^m_0 + \beta^m_1 p_i + \ep_i)] = \beta^m_0 \E\left[\hat{p}_i\right] + \beta^m_1 \E\left[\hat{p}_i p_i\right] + \E[\hat{p}_i \ep_i] \\ = \beta^m_0 \E[\hat{p}_i] + \beta^m_1 \omega \pi + \E[\hat{p}_i] \E[\ep_i] = \beta^m_0 \E[\hat{p}_i] + \beta^m_1 \omega \pi.
\end{multline*}
\item[5.] \textbf{Variance of $\hat{p}_i$}: Because $\hat{p}_i$ is binary, we have that $\V[\hat{p}_i] = \E[\hat{p}_i]\left(1 - \E[\hat{p}_i]\right) .$
\item[6.] \textbf{Covariance of $\hat{p}_i, m_i$}:
\begin{multline*}
\textrm{Cov}\left(\hat{p}_i, m_i\right) = \E\left[\hat{p}_i m_i\right] - \E[\hat{p}_i] \E[m_i] = \beta^m_0 \E[\hat{p}_i] + \beta^m_1 \omega \pi - \E[\hat{p}_i]( \beta^m_0 + \beta^m_1 \pi)\\ = \beta^m_1 \omega \pi - \E[\hat{p}_i] \beta_1^m \pi = \beta^m_1 \pi \left( \omega - \E[\hat{p}_i]\right).
\end{multline*}
Combining these expressions, we have that
$$ \hat{\beta}^m_1 \xrightarrow{P} \frac{\beta^m_1 \pi (\omega - \E[\hat{p}_i])}{\E[\hat{p}_i](1 - \E[\hat{p}_i])} = \beta^m_1 \gamma(\beta^g_0, \beta^g_1, c, \pi).$$
\end{itemize}
\subsubsection{Re-expressing $\gamma$ when $\pi = 1/2$}

Suppose $\pi = 1/2$. We rewrite the attenuation fraction $\gamma$ in a way that makes it more amenable to theoretical analysis.
